---
title: "Homework 2"
author: "jys2137"
date: "9/30/2021"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(tidyverse)
library(readxl)
```

The purpose of this file is to present the answers to Homework 2, an assignment reinforcing ideas in the **Data Wrangling I** topic for P8105. 


## Problem 0.1

This “problem” focuses on structure of my submission, especially the use git and GitHub for reproducibility, R Projects to organize my work, R Markdown to write reproducible reports, relative paths to load data from local files, and reasonable naming structures for my files.

To that end:

* I created a public GitHub repo + local R Project; we suggest naming this repo / directory p8105_hw2_jys2137
* I created a single .Rmd file named p8105_hw2_jys2137.Rmd that renders to github_document
* I created a subdirectory to store the local data files used in this assignment, and used relative paths to access these data files
* I will submit a link to my repo via Courseworks
* I aim to adhere to appropriate styling and clarity of code in Problems 1+.


## Problem 1

### Cleaning the _Mr. Trash Wheel_ sheet

The code chunk below is used to:

- **clean** the `Mr. Trash Wheel` sheet by:
  - omitting non-data entries (rows with notes / figures; columns containing notes)
  - using reasonable variable names (clean names)
  - omitting rows that do not include `dumpster`-specific data
- **round** `sports_balls` to the nearest integer

```{r clean_mr_trash_wheel}
mr_trash_wheel_sheet = 
  read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
    sheet = "Mr. Trash Wheel", skip = 1) %>%
  janitor::clean_names() %>%
  select(dumpster:homes_powered) %>%
  drop_na(dumpster) %>%
  mutate(sports_balls = round(sports_balls, digits = 0))
```

### Cleaning precipitation data for 2018 and 2019

The code chunk below **cleans** the precipitation data for 2018 and 2019 by:

- omitting rows without precipitation data 
- adding a variable for year

```{r clean_precip_2018_2019}
precip_2018 = 
  read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
    sheet = "2018 Precipitation", skip = 1) %>%
  janitor::clean_names() %>%
  drop_na(month) %>%
  mutate(year = 2018)

precip_2019 = 
  read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
    sheet = "2019 Precipitation", skip = 1) %>%
  janitor::clean_names() %>%
  drop_na(month, total) %>%
  mutate(year = 2019)
```

#### Combining precipitation datasets

Next, we **combine** the precipitation datasets and **convert month** to a character variable.

```{r combine_precip_data}
precip_comb_data = 
  bind_rows(precip_2018, precip_2019) %>% 
  mutate(
    month = month.name[as.numeric(month)], 
    month = str_to_lower(month))
```

### Description of the data

##### For the _Mr. Trash Wheel_ data

The _Mr. Trash Wheel_ dataset includes **`r nrow(mr_trash_wheel_sheet)`** observations, taking **`r ncol(mr_trash_wheel_sheet)`** variables into account. These `r ncol(mr_trash_wheel_sheet)` variables include information on each `dumpster` (including number, weight, and volume) and the amount of each litter type (including `plastic_bottles`, `polystyrene`, `cigarette_butts`, `glass_bottles`, `grocery_bags`, `chip_bags`, and `sports_balls`). The median number of sports balls in a dumpster in 2017 was **`r mr_trash_wheel_sheet %>% filter(year == 2017) %>% pull(sports_balls) %>% median()`**. 

##### For the _precipitation_ data

The precipitation data includes **`r nrow(precip_comb_data)`** observations, taking **`r ncol(precip_comb_data)`** variables into account. These **`r ncol(precip_comb_data)`** variables include `month`,  `total` precipitation in inches, and `year`. For available data, the total precipitation in 2018 was **`r precip_comb_data %>% filter(year == 2018) %>% pull(total) %>% sum()` in**.


## Problem 2

### Cleaning the _FiveThirtyEight_ datasets

This problem uses the _FiveThirtyEight_ datasets:`pols-month.csv`, `snp.csv`, and `unemployment.csv`. The following includes steps to clean and merge these into a single data frame using year and month as keys.

##### First, we clean `pols-month.csv`

```{r clean_pols}
pols_month_df = 
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv") %>%
  separate("mon", into = c("year", "month", "day"), convert = TRUE) %>% 
  mutate(
    month = month.name[as.numeric(month)], 
    month = str_to_lower(month),
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) %>% 
 select(-starts_with("prez"), -day)
```

##### Second, we clean `snp.csv`
For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.
```{r clean_snp}
snp_df = 
  read_csv("./data/fivethirtyeight_datasets/snp.csv") %>%
  separate("date", into = c("month", "day", "year"), convert = TRUE) %>% 
  arrange(year, month) %>% 
  mutate(
    month = month.name[(month)]) %>% 
  select(year, month, close, -day)
```

##### Third, we clean `unemployment.csv`

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r clean_unemployment}

```

##### Lastly, we merge the datasets

Join the datasets by merging snp into pols, and merging unemployment into the result.

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

Note: we could have used a date variable as a key instead of creating year and month keys; doing so would help with some kinds of plotting, and be a more accurate representation of the data. Date formats are tricky, though. For more information check out the lubridate package in the tidyverse.




