---
title: "Homework 2"
author: "jys2137"
date: "9/30/2021"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(tidyverse)
library(readxl)
library(knitr)

```

The purpose of this file is to present the answers to Homework 2, an assignment reinforcing ideas in the **Data Wrangling I** topic for P8105. 

## Problem 1

### Cleaning the _Mr. Trash Wheel_ sheet

The code chunk below is used to:

- **clean** the `Mr. Trash Wheel` sheet by:
  - omitting non-data entries (rows with notes / figures; columns containing notes)
  - using reasonable variable names (clean names)
  - omitting rows that do not include `dumpster`-specific data
- **round** `sports_balls` to the nearest integer

```{r clean_mr_trash_wheel}
mr_trash_wheel_sheet = 
  read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx",
    sheet = "Mr. Trash Wheel", skip = 1) %>%
  janitor::clean_names() %>%
  select(dumpster:homes_powered) %>%
  drop_na(dumpster) %>%
  mutate(sports_balls = round(sports_balls, digits = 0))
```

### Cleaning precipitation data for 2018 and 2019

The code chunk below **cleans** the precipitation data for 2018 and 2019 by:

- omitting rows without precipitation data 
- adding a variable for year

```{r clean_precip_2018_2019}
precip_2018 = 
  read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx",
    sheet = "2018 Precipitation", skip = 1) %>%
  janitor::clean_names() %>%
  drop_na(month) %>%
  mutate(year = 2018)

precip_2019 = 
  read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx",
    sheet = "2019 Precipitation", skip = 1) %>%
  janitor::clean_names() %>%
  drop_na(month, total) %>%
  mutate(year = 2019)
```

#### Combining precipitation datasets

Next, we **combine** the precipitation datasets and **convert month** to a character variable.

```{r combine_precip_data}
precip_comb_data = 
  bind_rows(precip_2018, precip_2019) %>% 
  mutate(
    month = month.name[as.numeric(month)], 
    month = str_to_lower(month))
```

### Description of the data

##### For the _Mr. Trash Wheel_ data

The _Mr. Trash Wheel_ dataset includes **`r nrow(mr_trash_wheel_sheet)`** observations, taking **`r ncol(mr_trash_wheel_sheet)`** variables into account. These `r ncol(mr_trash_wheel_sheet)` variables include information on each `dumpster` (including number, weight, and volume) and the amount of each litter type (including `plastic_bottles`, `polystyrene`, `cigarette_butts`, `glass_bottles`, `grocery_bags`, `chip_bags`, and `sports_balls`). The median number of sports balls in a dumpster in 2017 was **`r mr_trash_wheel_sheet %>% filter(year == 2017) %>% pull(sports_balls) %>% median()`**. 

##### For the _precipitation_ data

The precipitation data includes **`r nrow(precip_comb_data)`** observations, taking **`r ncol(precip_comb_data)`** variables into account. These **`r ncol(precip_comb_data)`** variables include `month`,  `total` precipitation in inches, and `year`. For available data, the total precipitation in 2018 was **`r precip_comb_data %>% filter(year == 2018) %>% pull(total) %>% sum()` in**.


## Problem 2

### Cleaning the _FiveThirtyEight_ datasets

This problem uses the _FiveThirtyEight_ datasets:`pols-month.csv`, `snp.csv`, and `unemployment.csv`. The following includes steps to clean and merge these into a single data frame using year and month as keys.

##### First, we clean `pols-month.csv`

We clean the data by:

- breaking up `mon` into integer variables `year`, `month`, and `day`
- replacing month number with month name
- creating `president` from values _gop_ and _dem_
- removing `prez_dem`, `prez_gop`, and `day`

```{r clean_pols}
pols_df = 
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv") %>%
  separate("mon", into = c("year", "month", "day")) %>% 
  mutate(
    month = month.name[as.numeric(month)], 
    month = str_to_lower(month),
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) %>% 
 select(-starts_with("prez"), -day)
```

##### Second, we clean `snp.csv`

For consistency across datasets, we arrange according to `year` and `month`, and organize so that `year` and `month` are the leading columns.

```{r clean_snp}
snp_df = 
  read_csv("./data/fivethirtyeight_datasets/snp.csv") %>%
  separate("date", into = c("month", "day", "year"), convert = TRUE) %>% 
  mutate(
    year, year = if_else(year <= 15, year + 2000, year + 1900)) %>% 
  arrange(year, month) %>% 
  mutate(
    month = month.name[as.numeric(month)], 
    month = str_to_lower(month),
    year = as.character(year)) %>% 
  select(year, month, close, -day)
```

##### Third, we clean `unemployment.csv`
Here, we tidy the unemployment data so that it can be merged with the previous datasets. This involves:

- switching from “wide” to “long” format
- ensuring that key variables have the same name
- ensuring that key variables take the same values

```{r clean_unemployment}
unemploy_df = 
  read.csv("./data/fivethirtyeight_datasets/unemployment.csv") %>% 
  janitor::clean_names() %>%
  pivot_longer(
    jan:dec,
    names_to = "month", 
    values_to = "unemployment") %>% 
  mutate(
      year = as.character(year),
      month = recode(month,
              jan = "january", feb = "february", mar = "march", apr = "april",
              may = "may", jun = "june", jul = "july", aug = "august",
              sep = "september", oct = "october", nov = "november", dec = "december"))
```

##### Lastly, we join the datasets

We join the datasets by merging snp into pols, and merging unemployment into the result.

```{r join_five_thirty_eight}
pols_snp_data =
  left_join(pols_df, snp_df)

fte_join_data =
  left_join(pols_snp_data, unemploy_df)
```

### Description of the _FiveThirtyEight_ datasets

The `pols_df` data includes **`r nrow(pols_df)`** observations and **`r ncol(pols_df)`** variables, related to the number of politicians who are democratic (`dem`) or republican (`gop`) at any given time between `r min(pols_df$year)` and `r max(pols_df$year)`.

The `snp_df` data includes **`r nrow(snp_df)`** observations and **`r ncol(snp_df)`** variables, related to Standard & Poor’s stock market index (S&P) from `r min(snp_df$year)` to `r max(snp_df$year)`. This included the `year`, `month`, and associated `close` values of the S&P.

The `unemploy_df` data includes **`r nrow(unemploy_df)`** observations and **`r ncol(unemploy_df)`** variables percentage of unemployment in each month (`january` through `december`) of the associated year from `r min(unemploy_df$year)` and `r max(unemploy_df$year)`.

The resulting dataset (with `pols_df`,`snp_df`, and `unemploy_df` combined) of **`r nrow(fte_join_data)`** observations and **`r ncol(fte_join_data)`** variables shows that between `r min(fte_join_data$year)` and `r max(fte_join_data$year)`, when a republican was `president`, the unemployment rate was **`r filter(fte_join_data, president == "gop") %>% pull(unemployment) %>% mean() %>% round(2)`%** and S&P was **`r filter(fte_join_data, president == "gop") %>% pull(close) %>% mean() %>% round(2)`** On the other hand, when a democrat was `president`, the `unemployment` rate was **`r fte_join_data %>% drop_na(unemployment) %>% filter(president == "dem") %>% pull(unemployment) %>% mean() %>% round(2)`%** and the S%P was **`r fte_join_data %>% drop_na(close) %>% filter(president == "dem") %>% pull(close) %>% mean() %>% round(2)`**

This dataset also has information on the `month` specific `unemployment` rates for each `year` and on the number of governors, senators, and representatives who were either democrats or republicans for each `month`.



## Problem 3
For this next section, we load and tidy the _Popular Baby Names_ dataset.

### Cleaning the _Popular Baby Names_ data

Here, we are using data from NYC Open data on the popularity of baby names. First, we tidy the data as shown in the code chunk below.

```{r baby_names}
baby_names_df = 
  read_csv("./data/Popular_Baby_Names.csv") %>% 
  janitor::clean_names() %>% 
  mutate(
    ethnicity = str_to_lower(ethnicity),
    childs_first_name = str_to_lower(childs_first_name),
    gender = str_to_lower(gender),
    ethnicity = recode(ethnicity, 
                       "asian and paci" = "asian and pacific islander", 
                       "black non hisp" = "black non hispanic",
                       "white non hisp" = "white non hispanic")) %>%
  distinct()

```

### Creating a table: rank in popularity of _"Olivia"_ as a female baby name 
Now, we produce a table showing the rank in popularity of the name “Olivia” as a female baby name over time. In this table: _ethnicities are in rows_ and  _years are in columns_.

```{r olivia_table}
olivia_table = 
  baby_names_df %>%
  filter(childs_first_name == "olivia") %>% 
  select(-gender, -childs_first_name, -count) %>% 
  arrange(year_of_birth) %>% 
  pivot_wider(names_from = year_of_birth, values_from = rank)

kable(olivia_table)

```

### Creating a table: most popular name among male children
Here, we produce a similar table as the one above, that shows the most popular name among male children over time. In this table: _ethnicities are in rows_ and  _years are in columns_.

```{r male_names}
male_names_table = 
  baby_names_df %>%
  filter(rank == "1", gender == "male") %>% 
  select(-gender, -count, -rank) %>% 
  arrange(year_of_birth) %>% 
  pivot_wider(names_from = year_of_birth, values_from = childs_first_name)

kable(male_names_table)

```

### Creating a scatter plot: rank of male white non-hispanic children names in 2016
Finally, we create a **scatter plot** for _male, white non-hispanic children born in 2016_. The number of children with a name (y axis) is against the rank in popularity of that name (x axis).
```{r final_scatterplot, warning=FALSE}
male_wnh_2016 = 
  baby_names_df %>% 
  filter(gender == "male", ethnicity == "white non hispanic", year_of_birth == 2016)

ggplot(male_wnh_2016, aes(x = rank, y = count, color = rank)) + 
  geom_point(alpha = .5) +
  scale_color_gradient(low = "blue", high = "red") +
  labs(
    title = "Popularity of names among male white non-hispanic children in 2016",
    x = "Rank in popularity",
    y = "Number of children",
    caption = "Data from Popular Baby Names dataset")

ggsave("male_wnh_2016.pdf", height = 4, width = 6)

```